NanoMaLy

### 2015-11-12 paper/pldi16-submission/semantics.tex
s/nat/int/g

Next, we formalize our search for witnesses to type-errors. We present
the syntax and dynamic semantics of \lang -- a simple lambda calculus with
integers and booleans, extended with our notion of holes -- as well as our
search algorithm. We prove that our system \emph{soundly} finds
witnesses, \ie if we find a witness then there is no possible typing for
the input program.

@@

\autoref{fig:syntax} describes the syntax of \lang, a simple lambda
calculus with numbers and booleans. As we are specifically interested in
programs that \emph{do} go wrong, we include an explicit \stuck state in
our syntax.

@@

\autoref{fig:operational} describes the small-step contextual reduction
semantics for \lang. The evaluation relation is parameterized by a pair
of \forcesym and \gensym functions that handle runtime type-checking and
instantiation of holes respectively. The relation must also maintain a
substitution \su, mapping holes to concrete values, so that we do not
instantiate the same hole with different values in different contexts,
and importantly, so that we can report a concrete witness to any
discovered type errors.

Note that each primitive reduction step -- addition, if-elimination, and
function application -- uses \forcesym to ensure that values have the
appropriate type (and that holes are instantiated) before continuing the
computation. Additionally, beta-reduction \emph{does not} type-check its
argument, it only ensures that the function can be applied.

\begin{thm}
\label{thm:all-reduce}
  Every closed expression $e$ reduces to a value $v$ (which may be \stuck).
\end{thm}
% \begin{proof}%[Proof of \autoref{thm:all-reduce}]
%   Simple induction on the evaluation relation.
% \end{proof}

\subsection{Soundness}
\label{sec:soundness}
We now show that our evaluation relation conservatively instantiates
holes, \ie if, given a function $f$, we report that 
$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su}$,
then for any type $t$ you assign to $f$, there exists an input \hastype{v}{t} such that
$\steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.

Intuitively, we can think of the evaluation of \eapp{f}{\ehole{}} as
computing a \emph{partial type} -- a type that may contain holes -- for
$f$. We can extract this type from the result of evaluation as follows.

>>>>

Next, we formalize our search for witnesses to type-errors. We present
the syntax and operational semantics of \lang -- a simple lambda calculus with
integers and booleans, extended with our notion of holes -- as well as our
search algorithm. We prove that our system \emph{soundly} finds
witnesses, \ie if we find a witness then there is no possible typing for
the input program.

@@

Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
calculus with numbers and booleans. As we are specifically interested in
programs that \emph{do} go wrong, we include an explicit \stuck state in
our syntax.

@@

Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang. The evaluation relation is parameterized
by a pair of \forcesym and \gensym functions that handle runtime
type-checking and instantiation of holes respectively. The relation must
also maintain a substitution \su, mapping holes to concrete values, so
that we do not instantiate the same hole with different values in
different contexts, and so that we can report a concrete witness to any
discovered type errors.

Note that each primitive reduction step -- addition, if-elimination, and
function application -- uses \forcesym to ensure that values have the
appropriate type (and that holes are instantiated) before continuing the
computation. Additionally, beta-reduction \emph{does not} type-check its
argument, it only ensures that the value being applied is a function.

\begin{thm}
\label{thm:all-reduce}
  Every closed expression $e$ reduces to a value $v$ (which may be \stuck).
  \ES{do we really need to state this, or is it obvious?}
\end{thm}
% \begin{proof}%[Proof of \autoref{thm:all-reduce}]
%   Simple induction on the evaluation relation.
% \end{proof}

\subsection{Soundness}
\label{sec:soundness}
We now show that our evaluation relation conservatively instantiates
holes, \ie if, given a function $f$, we report that 
$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su}$,
then for any type $\tfun{s}{t}$ you assign to $f$, there exists an input \hastype{v}{s} such that
$\steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.

We can think of the evaluation of \eapp{f}{\ehole{}} as
computing a \emph{partial type} -- a type that may contain holes -- for
$f$. We can extract this type from the result of evaluation as follows.
